---
title: "afscOM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{afscOM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The `afscOM` package can be downloaded and installed using: 

```{r setup}
remotes::install_github("BenWilliams-NOAA/afscOM")`
library(afscOM)
```

`afscOM` is an age structured, multi-sex, multi-fleet, spatially explicit simulation model, that can closely replicate the dynamics of a wide variety of stocks. 

Users can define the specific demographic parameters (natural mortality, maturity, weight-at-age, selectivity, etc.) that determine stock dynamics, as well as how those parameters vary across time, age, sex, region, and fleet. The package then provides a simple `project()` function that accepts an initial population state, the demographic parameters, and additional information about future recruitment and catch levels, and simulates the population forward in time. 

Users can, optionally, also generate observations (with provided levels of observation error) from the population.

#### Test Cases
`afscOM` was designed to be used with a management strategy evaluation (MSE) framework for evaluating possible alternative management strategies for Alaska sablefish (_Anoplopoma fimbria_). As such it has been rigorously tested for its ability to reproduce the estimated timeseries of sablefish biomass, fishing mortality, and catches (Goethel et al. 2023), as well as its ability to reproduce survey observations. Additionally, spatially-explicit features have been compared to the resulted achieved by the eastern-Pacific coastwide sablefish model (Kapur 2023, Kapur et al. 2024), and an spatially-explicit version of the Alaska sablefish model (Cheng et al. 2025).

If you would like to test `afscOM` for its ability to recreate the dynamics of your own stock, please contact the project maintainers.

#### References

Goethel, D.R., Cheng, M.L.H., Echave, K.B., Marsh, C., Rodgveller, C.J., Shotwell, K., Siwicke, K., 2023. Stock Assessment of the sablefish stock in Alaska. North Pacific Fisheries Management Council, Anchorage, AK.

Kapur, M.S. 2023. A Management Strategy Evaluation for Transboundary Sablefish in the Northeast Pacific Ocean. Dissertation, University of Washington, Seattle, WA.

Kapur, M.S., Haltuch, M.A., Connors, B.M., Berger, A.M., Holt, K., Marshall, K.N., and Punt, A.E. 2024. Range-wide contrast in management outcomes for transboundary Northeast Pacific sablefish. Can. J. Fish. Aquat. Sci. NRC Research Press. doi:10.1139/cjfas-2024-0008.

Cheng, M.L.H., Marsh, C.A., Goethel, D.R., Hulson, P.-J.F., Echave, K., Williams, B.C., Berger, A.M. and Cunningham, C.J. (2025), Panmictic Panacea? Demonstrating Good Practices for Developing Spatial Stock Assessments Through Application to Alaska Sablefish (Anoplopoma fimbria). Fish Fish, 26: 825-847. https://doi.org/10.1111/faf.70002