---
title: "Model Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A vareity of model options also need to be specified to ensure the projection model works as expected. Required options are:

* `removals_input`: input units of removals (either "catch" is inputs are a TAC, or "F" if an instantenous fishing mortality rate)
* `simulate_observations`: whethr to simulate observations from the OM (TRUE/FALSE)

```{r, base_model_options}
model_options <- list(
  removals_input = "catch",
  simulate_observations = TRUE
)
```

### Catch Apportionment
If `removals_input="catch"`, then region and fleet-level apportionment must also be specified: 

* `region_apportionment`: an unnamed list of vectors (of length `nyears`) specifying what proportion of the TAC should be allocated to each spatial region. There should be as many vector as there are regions. The sum across all regions in a given year should be 1.0, but the model will not make this check internally.

* `fleet_apportionment`: an unnamed list of vectors (of length `nyears`) specifying what proportion of the TAC should be allocated to each fishing fleet. There should be as many vectors as there are fishing fleets. For multi-region models, the `fleet_apportionment` proportions will be applied across all regions. There is not currently support for specifying unique fleet apportionments per spatial region. 

```{r apportionment_options, eval=FALSE}
# For a single region model, 100% of the catch is apportioned to the single region
model_options$regional_apportionment <- list(1)

# Define a 70%-30% apportionment split between fleet 1 and fleet 2
model_options$fleet_apportionment <- list(rep(0.70, nyears), rep(30, nyears))

```

### Observation Processes
If `simulate_observations=TRUE`, then a set of parameters that govern how observations are simulated must also be defined. This option, `obs_pars` takes the form of a complex list, where each named element of the list refers to a specific survey or fishery from which observations are being simulated. Currently, there is support for a longline survey (`surv_ll`), a trawl survy (`surv_tw`), a fixed gear fishery (`fish_fx`), and a trawl fishery (`fish_tw`). 

Within each survey/fishery, specific parameter values that define the observation processes must be defined. The parameters supported for survey and fishery fleets differ somewhat.

For surveys, the following observation parameters must all be defined:

* `q`: a catchability coefficient
* `rpn_cv`: coefficient of variation of RPN observations
* `rpw_cv`: coefficient of variation of RPW observations
* `ac_samps`: the number of samples, across all sexes, from which age composition data is simulated
* `as_integers`: whether to return age composition observations as integer or proportions

For fishery fleets, only the following observation parameters need to be defined:

* `ac_samps`: the number of samples, across all sexes, from which age composition data is simulated
* `as_integers`: whether to return age composition observations as integer or proportions

For additional information on how all of these parameters are used to generate observations, see ["Observation Processes"](observation_processes.html).

```{r, obs_pars}
obs_pars <- list(
    surv_ll = list(
        q = 6.41338,
        rpn_cv = 0.20,
        rpw_cv = 0.10,
        ac_samps = 1000,
        as_integers = TRUE
    ),
    surv_tw = list(
        q = 0.8580,
        rpw_cv = 0.10,
        ac_samps = 1000,
        as_integers = TRUE
    ),
    fish_fx = list(
        ac_samps = 1000,
        as_integers = TRUE
    ),
    fish_tw = list(
        ac_samps = 1000,
        as_integers = TRUE
    )
)

model_options$obs_pars <- obs_pars
```

### Default Options
A helper function, `setup_model_options()` will return a fully formed `model_options` list object with required parameters set to default values.

The default values are as follows:

* `removals_input`: "catch"
* `simulate_observations`: TRUE
* `region_apportionment`: 1/nregions
* `fleet_apportionment`: 1/nfleets
* `*$q`: 1
* `*$rpn_cv`: 0.10
* `*$rpw_cv`: 0.10
* `*ac_samps`: 100
* `*as_integers`: TRUE